---
title: "Severe Maternal Morbidity"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: "https://github.com/chokatie/data555_final/blob/main/Part4a.Rmd"
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(flexdashboard)
library(shiny)
library(haven)
library(tidyverse)
library(plotly)
library(RColorBrewer)
```

```{r data-prep}
thesis_data <- read_sas("clean_categsv2.sas7bdat")

# Label MOTHER_RACE codes with actual corresponding race names
thesis_data <- thesis_data |>
  mutate(
    race = factor(
      MOTHER_RACE,
      levels = c(1, 2, 3, 4, 5, 6),
      labels = c("White",
                 "Black",
                 "Asian",
                 "American Indian",
                 "Pacific Islander",
                 "Hispanic")
    )
  )

# Race-level summary: percent SMM and rate per 10,000
smm_race <- thesis_data |>
  group_by(race) |>
  summarise(
    smm_percent = mean(comp_smm, na.rm = TRUE) * 100,
    smm_rate_10k = mean(comp_smm, na.rm = TRUE) * 10000,
    .groups = "drop"
  )

# Identify the White reference
white_rate <- smm_race |>
  filter(race == "White") |>
  pull(smm_percent)

smm_race <- smm_race |>
  mutate(
    pct_diff_white = 100 * (smm_percent - white_rate) / white_rate,
    hover_text = paste0(
      "Race: ", race, "<br>",
      "SMM risk: ", round(smm_percent, 1), "%<br>",
      "SMM rate: ", round(smm_rate_10k, 1), " per 10,000 deliveries<br>",
      "% difference vs White: ", round(pct_diff_white, 1), "%"
    )
  )

# Summary data for Widget 2
smm_time <- thesis_data |>
  group_by(year, race) |>
  summarise(
    smm_percent = mean(comp_smm, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  mutate(
    hover_text = paste0(
      "Year: ", year, "<br>",
      "Race: ", race, "<br>",
      "SMM risk: ", round(smm_percent, 1), "%"
    )
  )

yrs   <- sort(unique(smm_time$year))
races <- levels(thesis_data$race)
```

**Dashboard** {data-icon="fa-chart-line"}
=========================================

Row
-------------------------------------

### **Overview**
>  1. How do SMM risk rates differ across major US racial groups?  
  2. How have SMM risks for these groups changed over time (2016–2022)?  

*This dashboard compares severe maternal morbidity (SMM) risk across major U.S. racial groups and shows how those risks changed from 2016 to 2022. Use the interactive charts below to explore these questions.* <br><br>

Row {data-height=380}
-------------------------------------

### Figure 1 {data-width=70}

```{r bar-chart}
bar1 <- brewer.pal(n = length(unique(smm_race$race)), name = "Dark2")

plot_ly(
  smm_race,
  x = ~race,
  y = ~smm_percent,
  type = "bar",
  hovertext = ~hover_text,
  hoverinfo = "text",
  marker = list(color = bar1)
) |>
  layout(
    title = list(
      text = "SMM Risk Rates across US Racial Groups",
      x = 0.5,
      xanchor = "center",
      font = list(
        family = "Arial Bold",
        size = 20,
        color = "black"
      )
    ),
    height = 350,
    xaxis = list(
      title = list(
        text = "Race",
        font = list(
          family = "Arial Bold",
          size = 16,
          color = "black"
        )
      ),
      automargin = TRUE,
      tickfont = list(
        family = "Arial",
        size = 12,
        color = "black"
      )
    ),
    yaxis = list(
      title = list(
        text = "Risk (%)",
        font = list(
          family = "Arial Bold",
          size = 16,
          color = "black"
        )
      ),
      tickfont = list(
        family = "Arial",
        size = 12,
        color = "black"
      )
    ),
    margin = list(t = 85, b = 85, l = 75, r = 20),
    hoverlabel = list(
      bgcolor = "white",
      bordercolor = "black",
      font = list(
        family = "Arial",
        size = 12,
        color = "black"
      )
    )
  ) |>
  config(displayModeBar = TRUE)
```

### **Key Findings** {data-width=30}

- White and Asian women have the lowest SMM risk prevalence in this dataset.

- Black and American Indian women have the highest SMM risk prevalence.


Row
-------------------------------------

### Figure 2 {data-width=75}

```{r trend-chart}
renderPlotly({
  df <- smm_time |>
    filter(
      year >= input$year_range[1],
      year <= input$year_range[2],
      race %in% input$race_sel
    )

  plot_ly(
    df,
    x = ~year,
    y = ~smm_percent,
    color = ~race,
    type  = "scatter",
    mode  = "lines+markers",
    text  = ~hover_text,
    hoverinfo = "text"
  ) |>
    layout(
      title = list(
        text = "Time Trends in SMM Risk, 2016–2022",
        x = 0.5,
        xanchor = "center",
        font = list(
          family = "Arial Bold",
          size = 20,
          color = "black"
        )
      ),
      xaxis = list(
        title = list(
          text = "Year",
          font = list(
            family = "Arial Bold",
            size = 16,
            color = "black"
          )
        ),
        dtick = 1,
        tickfont = list(
          family = "Arial",
          size = 12,
          color = "black"
        )
      ),
      yaxis = list(
        title = list(
          text = "Risk (%)",
          font = list(
            family = "Arial Bold",
            size = 16,
            color = "black"
          )
        ),
        tickfont = list(
          family = "Arial",
          size = 12,
          color = "black"
        )
      ),
      legend = list(
        orientation = "h",
        x = 0,
        y = -0.2,
        font = list(size = 11)
      ),
      margin = list(t = 85, b = 95, l = 75, r = 20)
    )
})
```

### Filters {data-width=25}

```{r filters}
sliderInput(
  "year_range", "Year range:",
  min   = min(yrs),
  max   = max(yrs),
  value = c(min(yrs), max(yrs)),
  step  = 1,
  sep = "",
  width = "100%"
)

checkboxGroupInput(
  "race_sel", "Race(s):",
  choices  = races,
  selected = c("Asian", "White"),
  inline   = FALSE
)
```

Row
-------------------------------------

### **Key Findings**

- Among the selected groups, SMM risk generally increases over time from 2016 to 2022.

- Asian women consistently have slightly higher SMM risk than White women over 2017–2022 (2016 was the only year where Asian women had lower SMM risk prevalence), however, the gap between the two groups remains relatively small and stable across years.
<br><br>


Row
-------------------------------------

### **Why Does This Matter?** {data-width=70}

Severe maternal morbidity is an important public health outcome as it reflects serious pregnancy-related complications and can reveal inequities in maternal care. This dashboard helps identify racial disparities and time trends that may inform maternal health policy, equity-focused interventions, and quality improvement efforts.
<br><br>

### *Project Links* {data-width=30}

- [GitHub repository](https://github.com/chokatie/data555_final)
- Source code linked in Github repo & Dashboard *(upper right corner)* 

**Data Description** {data-icon="fa-database"}
==============================================

Row
-------------------------------------

### **About the Dataset**
#### *Data Source*

> Data were obtained from the Georgia Department of Public Health (GA-DPH) and included full-year consolidated vital statistics records and hospital discharge files from 2016-2022 (initial n=918,752). The vital records, consisting of live birth and fetal death certificates, were then deterministically linked with statewide inpatient discharge data using identifiers maintained by the GA-DPH.


#### *Data Components*

> From these linked files, various exclusionary criteria were applied to omit deliveries at facilities with fewer than 10 births, missing maternal race, missing data on any study covariate to result in an analytic sample of 675,330 deliveries. Severe maternal morbidity was measured using a composite indicator called `comp_smm`, and race was based on the coded maternal race variable in the source dataset. 

#### *Privacy*

> The data were assembled from existing administrative and clinical sources in partnership with the Georgia Department of Public Health and Emory University, thus the dataset is confidential and the raw data are not posted publicly.